---
title: "Course_Project_Part_2"
author: "suveyda"
date: "February 2, 2017"
output: html_document
---

# Course Project Part 2

## Statistical Inference by Johns Hopkins University

### Introduction

Now in the second part of the project, we will analyze the ToothGrowth data in the R datasets package. Using this dataset, our goal is to investigate if any of the variables have an impact on the tooth growth.

```{r, include=FALSE}
library(datasets)
library(ggplot2)
library(gridExtra)
```

### Data

The ToothGrowth data in the R datasets package come from a research which is focused on finding the Effect of Vitamin C on Tooth Growth in Guinea Pigs. The data set consists of the results of 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC). The response is the length of odontoblasts (cells responsible for tooth growth).

Dataset contains 60 observations on 3 variables:

*	**len:**	numeric	Tooth length
*	**supp:**	factor	Supplement type (VC or OJ).
*	**dose:**	numeric	Dose in milligrams/day

So, we first load the ToothGrowth data to R, and take a look at the variables:

```{r}
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
str(ToothGrowth)
```

### Exploratory Data Analyses

At this step, we perform some basic exploratory data analyses:

```{r}
summary(ToothGrowth)
```

As you can see in the below table, all possible combinations of "supp" and "dose" have same number of samples:

```{r}
table(ToothGrowth$supp, ToothGrowth$dose)
```

Now let's compare the variances between two supplement groups:

```{r}
var(ToothGrowth$len[ToothGrowth$supp == 'OJ'])
var(ToothGrowth$len[ToothGrowth$supp == 'VC'])
```

And now the variances of three dose levels:

```{r}
var(ToothGrowth$len[ToothGrowth$dose == 0.5])
var(ToothGrowth$len[ToothGrowth$dose == 1])
var(ToothGrowth$len[ToothGrowth$dose == 2])
```

Based on the above values, we will assume unequal variances in the statistical analysis. 

Furthermore, below we have the violin plots of tooth length data by supplement type and dose. As you can see, supplement type does not have a significant impact on the toothgrowth while dose level has an positive effect on the toothgrowth. This is just our initial estimate, let's take a detail look to the two factors before starting the statistical analysis. 

```{r violin.plots}
supp.plot <- ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp)) + geom_violin(col = "black") + labs(x = "Supplement Type", y = "Tooth Length")

dose.plot <- ggplot(ToothGrowth, aes(x = dose, y = len, fill = dose)) + geom_violin(col = "black") + labs(x = "Dose (mg/day)", y= "Tooth Length")

grid.arrange(supp.plot, dose.plot, ncol = 2, top = "Violin Plots of Tooth Length Data by Supplement Type and Vitamin C Dose")
```

In the below box plots, we can see effect of vitamin dose on tooth length separatelly for two supplement types. According to the plots, in both supplement types, vitamin dose has an positive effect on the tooth growth, but we will see if that is in fact the case in the next part.

```{r dose.box.plot}
dose.box.plot <- ggplot(ToothGrowth, aes(x = dose, y = len)) + facet_grid(.~supp) + geom_boxplot(aes(fill = dose)) + labs(x = "Dose (mg/day)", y = "Tooth Length")
dose.box.plot
```

Similarly, in the below box plots, we can see effect of supplement type on tooth length separatelly for three vitamin dosages. According to the plots, when the vitamin dose is 0.5 or 1 mg per day, orange juice is a better supplement for tooth growth than ascorbic acid. On the other hand, when the vitamin dose is 2 mp per day, supplement type has no effect on the tooth growth. Let's see if we can prove this in the next section statistically. 

```{r supp.box.plot}
supp.box.plot <- ggplot(ToothGrowth, aes(x = supp, y = len)) + facet_grid(.~dose) + geom_boxplot(aes(fill = supp)) + labs(x = "Supplement Type", y = "Tooth Length")
supp.box.plot
```

### Statistical Inference Analysis

In this part of the project, we will investigate the relationship between tooth growth vs. supplement type and vitamin dose. We will first investigate the effect of supplement type and vitamin dose on tooth growth separatelly, and then analyse their effect together.

#### Effect of supplement type on tooth growth

**Hypothesis:**

* H~0~: There is **no** difference in tooth growth when using orange juice or ascorbic acid as the supplement type
* H~a~: There is a difference in tooth growth when using orange juice or ascorbic acid as the supplement type

```{r}
OJ = ToothGrowth$len[ToothGrowth$supp == 'OJ']
VC = ToothGrowth$len[ToothGrowth$supp == 'VC']

t.test(OJ, VC)
```

Here, p-value is larger than 0.05 and the confidence interval includes zero. Therefore, we cannot reject the null hypothesis, and conclude that supplement type has no effect on tooth growth.

#### Effect of vitamin dose on tooth growth

**Hypothesis:**

* H~0~: There is **no** difference in tooth growth when receiving different levels of vitamin C dosage
* H~a~: There is a difference in tooth growth when receiving different levels of vitamin C dosage

```{r}
half = ToothGrowth$len[ToothGrowth$dose == 0.5]
one = ToothGrowth$len[ToothGrowth$dose == 1]
two = ToothGrowth$len[ToothGrowth$dose == 2]
```

**0.5 vs 1 mg/day**

```{r}
t.test(half, one)
```

According to the t-test, p-value is less than 0.05 and interval is entirely below zero which means that we can reject the null hypothesis and conclude that there is a difference in tooth growth when receiving different levels of vitamin C dosage (0.5 mg/day vs 1 mg/day).

**0.5 vs 2 mg/day**

```{r}
t.test(half, two)
```

Similarly, the p-value is less than 0.05 and interval is entirely below zero which suggest that receiving 0.5 mg vitamin per day leads to less growth in tooth length than receiving 2 mg vitamin per day.

**1 vs 2 mg/day**

```{r}
t.test(one, two)
```

Finally, when we compare 1 mg/day and 2 mg/day, the p-value is less than 0.05 and interval is entirely below zero. Therefore, we can conclude that receiving 2 mg vitamin per day cause more tooth growth than receiving 0.5 mg vitamin per day.

In conclusion, we see that there is a positive correlation between vitamin dose and tooth growth. So, as we increase the vitamin dose, we will observe larger growth in tooth length. 

#### Effect of vitamin dose and supplement type on tooth growth

**Hypothesis:**

* H~0~: There is **no** difference in tooth growth when receiving different levels of vitamin C dosage through orange juice or ascorbic acid as the supplement type
* H~a~: There is a difference in tooth growth when receiving different levels of vitamin C dosage through orange juice or ascorbic acid as the supplement type

Here we will first compare different supplement types when the vitamin dose in constant:

```{r}
test.results.1 = data.frame(comparisons = character(), pvalues = numeric(), interval.start = numeric(), interval.end = numeric())

t <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 0.5))
test.results.1 <- rbind(test.results.1, data.frame(comparisons = "0.5 mg with OJ vs 0.5 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 1))
test.results.1 <- rbind(test.results.1, data.frame(comparisons = "1 mg with OJ vs 1 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ supp, data = subset(ToothGrowth, dose == 2))
test.results.1 <- rbind(test.results.1, data.frame(comparisons = "2 mg with OJ vs 2 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

test.results.1
```

As you can see from the above table, for 0.5 mg/day and 1 mg/day dosages, p-values are both below 0.05 and confidence intervals are both above zero. Therefore, we can reject the null hypothesis and conclude that when either one of these two dose levels are received, orange juice leads to higher tooth growth than ascorbic acid. However, when the vitamin dose is 2 mg/day, p-value is larger than 0.05 and the interval contains zero. So, we cannot reject the null hypothesis.

Now we will compare effects of receiving different vitamin dose levels when the supplement type is constant:

```{r}
test.results.2 = data.frame(comparisons = character(), pvalues = numeric(), interval.start = numeric(), interval.end = numeric())

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "OJ" & dose != 2))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "0.5 mg with OJ vs 1 mg with OJ", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "OJ" & dose != 1))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "0.5 mg with OJ vs 2 mg with OJ", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "OJ" & dose != 0.5))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "1 mg with OJ vs 2 mg with OJ", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "VC" & dose != 2))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "0.5 mg with VC vs 1 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "VC" & dose != 1))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "0.5 mg with VC vs 2 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

t <- t.test(len ~ dose, data = subset(ToothGrowth, supp == "VC" & dose != 0.5))
test.results.2 <- rbind(test.results.2, data.frame(comparisons = "1 mg with VC vs 2 mg with VC", pvalues = t$p.value, interval.start = t$conf.int[1], interval.end = t$conf.int[2]))

test.results.2
```

In the above table, all p-values are below 0.05 and all intervals are below 0. Therefore, for all combinations of vitamin dose and supplement type, we can reject the null hypothesis. Furthermore, since all intervals are below 0, we can conclude as we increase the vitamin dose we will observe increase in tooth growth.

### Conclusions

Before summarizing the conclusions of this analysis, let's first list our assumptions:

* Underlying data is independent and identically distributed (i.i.d.)
* Groups have unequal variances
* Distribution of the means is normal
* Sample is a good representation of the population

Finally, let's summarize the conclusions of this analysis:

* In general, supplement type has no effect on tooth growth. However, orange juice leads to higher tooth growth than ascorbic acid when pigs receive either 0.5 mg or 1 mg Vitamin C daily.

* There is a positive correlation between vitamin dose and tooth growth. So, as we increase the vitamin dose, we will observe higher tooth growth.
